---
title: "Exercise 1"
author: ""
date: "2023-04-07"
description: >
  Questions for material for up to Module 3.
output: 
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Exercise 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  chunk = hook_foldable("chunk")
  #output = hook_foldable("output"),
  #plot = hook_foldable("plot")
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  #class.source = 'fold-hide',
  #foldcode=TRUE
)
```

## Question 1

Install and load library `dplyr`.

```{r eval=FALSE}
# install package
install.packages(dplyr)
```

```{r}
# load dplyr package
library(dplyr)
```

a.  Call help for function arrange of `dplyr`

```{r}
# help page of dplyr
?dplyr
```


\

## Question 2

Create a variable called `myVar` and assign it to $30$

```{r}
myVar <- 30
```

a.  What data type is `myVar`? (Use a function to check)

```{r}
class(myVar)
typeof(myVar)
```

b.  Convert `myVar` to a different data type

```{r}
# convert to string
myVar <- as.character(myVar)
class(myVar)

# convert to integer
myVar <- as.integer(myVar)
class(myVar)
```


\

## Question 3

Create a new variable called `my_num` that contains 6 numbers

```{r}
my_num <- 16:21
my_num
```

a.  Multiply `my_num` by 4

```{r}
my_num * 4
```


\

## Question 4

Create a second variable called `my_char` that contains 5 character strings

```{r}
my_char <- c("nn", "oo", "rr", "ee", "yy")
```


\

## Question 5

Combine the two variables that `my_num` and `my_char` into a variable called `both`

```{r}
both <- c(my_num, my_char)
both
```


a.  What is the length of `both`

```{r}
length(both)
```

b.  Divide `both` by 3, what happens?

```{r error=TRUE}
both/3
```



\

## Question 6

Create a vector with elements 1, 2, 3, 4, 5, 6 and call it `x`

```{r}
x <- 1:6
#x <- c(1, 2, 3, 4, 5, 6)
x
```


\

## Question 7

Create another vector with elements 10, 20, 30, 40, 50 and call it `y`

```{r}
y <- seq(from = 10, to = 50, by = 10)
#y <- c(10, 20, 30, 40, 50)
y
```


\

## Question 8

What happens if you try to add `x` and `y` together? Why?

a.  Add `x` and `y` together

```{r}
x + y

##
# We notice that the two variables have different lengths, such that the sixth value of `x` is added to the first value in `y`. We also receive a warning for this operation.
```

b.  Multiply `x` and `y` together. Pay attention to how R performs operations on vectors of the same length

```{r}
x * y
```


c.  What happens if `x` and `y` have the same length?

```{r}
x <- 1:5
x + y
x * y 

```


\

## Question 9

Now create a vector with elements "aa", "bb", "cc", "dd", "kk" and call it `z`

```{r}
z <- c("aa", "bb", "cc", "dd", "kk") 
```


a.  What happens if you try to add `x` or `y` with `z` together

```{r error=TRUE}
x + z
```

b.  Why does the error comes up? (Hint: data types)

```{r}
##
# We notice that the data type of `x` and `y` (numeric), is different from `z` (character), and we are not able to perform computational operations on different data types.
```


\

## Question 10

If `x <- c("ww", "ee", "ff", "uu", "kk")`, what is the output for `x[c(2,3)]`?

```{r}
x <- c("ww", "ee", "ff", "uu", "kk")
x[c(2, 3)]
```

a.  Can you subset `x` in another way and get the same output?

```{r}
x[2:3]
```


\

## Question 11

If `x <- c("ss", "aa", "ff", "kk", "bb")`, what will be the third value in the index vector operation `x[c(2,4,4)]`?

```{r}
x <- c("ss", "aa", "ff", "kk", "bb")
x[c(2,4,4)]

##
# The third value is "kk", which is the fourth value in `x`
```


\

## Question 12

Assign `fname <-  "James"` and `lname <- "Bond"`,

```{r}
fname <-  "James"
lname <- "Bond"
```

a.  Produce the output `James Bond` using a line of code (Hint: use `paste()`)

```{r}
paste(fname, lname)
# paste0(fname, " ", lname)
```


\

## Question 13

Create two character vectors:

-   `Names <- c("John", "Andrew", "Thomas")` and
-   `Designation <- c("Manager", "Project Head", "Marketing Head")`.

a.  Write some R code to obtain the following output (Hint: check help page the function you used to create a data frame):

```{r echo=FALSE, fold.chunk=FALSE}
Names <- c("John", "Andrew", "Thomas")
Designation <- c("Manager", "Project Head", "Marketing Head")

names_design <- paste(Names, Designation)
df <- data.frame("Names Designation" = names_design, check.names = FALSE)
df
```


```{r eval=FALSE}
Names <- c("John", "Andrew", "Thomas")
Designation <- c("Manager", "Project Head", "Marketing Head")

names_design <- paste(Names, Designation)
df <- data.frame("Names Designation" = names_design, check.names = FALSE)
df
```


\

## Question 14

Create the following data frame, and call it Employees

```{r echo=FALSE, fold.chunk=FALSE}
Names <- c("Alex", "Lilly", "Mark", "Oliver", "Martha", "Lucas", "Caroline")
Age <- c(25, 31, 23, 52, 76, 49, 26)
Height <- c(177, 163, 190, 179, 163, 183, 164)
Weight <- c(57, 69, 83, 75, 70, 83, 53)
Sex <- c("F", "F", "M", "M", "F", "M", "F")

Employees <- data.frame(Names, Age, Height, Weight, Sex)
Employees
```


```{r eval=TRUE}
Names <- c("Alex", "Lilly", "Mark", "Oliver", "Martha", "Lucas", "Caroline")
Age <- c(25, 31, 23, 52, 76, 49, 26)
Height <- c(177, 163, 190, 179, 163, 183, 164)
Weight <- c(57, 69, 83, 75, 70, 83, 53)
Sex <- c("F", "F", "M", "M", "F", "M", "F")

Employees <- data.frame(Names, Age, Height, Weight, Sex)
Employees
```

a.  Extract specific column from a data frame

```{r}
# extract the height column form the data frame
Employees$Height
```

b.  Extract 3rd and 5th rows with 1st and 3rd columns

```{r}
Employees[c(3, 5), c(1, 3)]
```

c.  Add a new column and call it Education, where everyone has an education status of HS

```{r}
Employees$Education <- "HS"
Employees
```

d.  Add new row called Louis

```{r}
louis_row <- c("Louis", 34, 185, 80, "M", "HS")
Employees <- rbind(Employees, louis_row)
Employees
```

e.  Make `Sex` column into a factor. How many factor levels are in the columns?

```{r}
Employees$Sex <- factor(Employees$Sex)
#Employees <- mutate(Employees, Sex = factor(Sex))
class(Employees$Sex)
```


\

## Question 15

Create the data frame below and call it Working (make sure you import the variable Working as character and not factor).

```{r echo=FALSE}
Names <- c("Alex", "Lilly", "Mark", "Oliver", "Martha", "Lucas", "Caroline", "Louis")
Working <- c("Yes", "No", "No", "Yes", "Yes", "No", "Yes", "Yes")
Working <- data.frame(Names, Working)
Working
```

```{r eval=FALSE}
Names <- c("Alex", "Lilly", "Mark", "Oliver", "Martha", "Lucas", "Caroline", "Louis")
Working <- c("Yes", "No", "No", "Yes", "Yes", "No", "Yes", "Yes")
Working <- data.frame(Names, Working)
Working
```

a.  Add this data frame column-wise to the previous one.

```{r}
# column bind the new column 
Employees_new <- cbind(Employees, Working = Working$Working)

## alternative:
#Employees_new <- dplyr::left_join(x = Employees, y = Working, by = "Names")

Employees_new
```

b.  How many rows and columns does the new data frame have?

```{r}
dim(Employees_new)

##
# The new data frame has 8 rows and 7 columns.
```

c.  What class of data is in each column?

```{r}
str(Employees_new)

##
# `Names`, `Age`, `Height`, `Weight`, `Education`, `Working` are character (string), and `Sex` is a factor.
```

d. Convert the `Age`, `Height` and `Weight` to numeric data type or integer data type using `as.[data type]()` (Hint: `as.numeric()`) ; convert `Education` and `Working` to factor variable using `dplyr::mutate()`.

```{r}
# convert to numeric data type
Employees_new$Age <- as.numeric(Employees$Age)
Employees_new$Height <- as.numeric(Employees_new$Height)

# convert to integer data type
Employees_new$Weight <- as.integer(Employees_new$Weight)

# convert to factor variable
Employees_new <- dplyr::mutate(Employees_new,
                               Education = factor(Education),
                               Working = factor(Working))

##
# Or all together
# Employees_new <- Employees_new |> 
#   dplyr::mutate(
#     Age = as.numeric(Age),
#     Height = as.numeric(Height),
#     Weight = as.integer(Weight),
#     Education = factor(Education),
#     Working = factor(Working)
#   )

str(Employees_new)
```



\

# **Using the `iris` dataset in R**

## Question 16

Extract `Sepal.Length` from the `iris` dataset and call the resulting vector `mysepal`

```{r}
mysepal <- iris$Sepal.Length
head(mysepal)
```


a.  Get the summation, mean, median, max and min of `mysepal`

```{r}
# summation
sum(mysepal)

# mean
mean(mysepal)

# median
median(mysepal)

# max
max(mysepal)

# min
min(mysepal)
```

b.  Get the summary of `mysepal`

```{r}
summary(mysepal)
```


\

## Question 17

Get all rows of Species `versicolor` in a new data frame. Call this data frame: `iris.vers`

```{r}
iris.vers <- subset(iris, Species == "versicolor")
#iris.vers <- iris[iris$Species == "versicolor", ]
iris.vers
```


\

## Question 18

Get a vector called `sepal.dif` with the difference between `Sepal.Length` and `Sepal.Width` of `versicolor` plants

```{r}
sepal.dif <- iris.vers$Sepal.Length - iris.vers$Sepal.Width
sepal.dif
```


\

## Question 19

Update (add) `iris.vers` with the new column `sepal.dif`

```{r}
iris.vers$`Sepal_diff` <- sepal.dif
#iris.vers <- cbind(iris.vers, Speal_diff = sepal.dif)
head(iris.vers)
```


\

## Question 20

How many observations in `iris` gave a Sepal Length greater or equal 5?

```{r}
# nrow() finds the number of rows (observations)
# subset() select the Sepal.Length >= 5
nrow(subset(iris, Sepal.Length >= 5))

##
# alternatively:
# iris |> 
#   dplyr::filter(Sepal.Length >= 5) |> 
#   nrow()
```


\

# **Using the `mtcars` dataset in R**

## Question 21

Check to see if you have the `mtcars` dataset by entering the command `mtcars`.

a.  What class is `mtcars`?

```{r}
class(mtcars)

##
# `mtcars` is a data frame
```


\

## Question 22

How many variables (columns) and observations (rows)are in the `mtcars` dataset?

```{r}
dim(mtcars)

##
# There are 32 observations and 11 variables in the `mtcars` dataset.
```


\

## Question 23

Perform an descriptive analysis on the `mtcars` dataset (Hint: data types of columns, mean, median, ...).

```{r}
# find internal structure
str(mtcars)

# find the data distribution of the `mtcars` dataset
summary(mtcars)

# find the correlation between the `mtcars` variables
cor(mtcars)
```

a.    What are the unique values in `cyl`, and `vs` (Hint: use `unique()`).

```{r}
# unique values in cyl
unique(mtcars$cyl)

# unique values in vs
unique(mtcars$vs)
```


b.    How many unique values in `am` and `gear` (Hint: use `length()`).

```{r}
# number of unique values in am
length(unique(mtcars$am))

# number of unique values in gear
length(unique(mtcars$gear))
```



c.    Bonus: Try using `sapply()` to find the unique values and the number of unique values in `cyl`, `vs`, `am`, `gear`, and `carb` all at once.

```{r}
# for each variable
length(unique(mtcars$cyl))
length(unique(mtcars$vs))
```

```{r}
sapply(c("cyl", "vs", "am", "gear", "carb"), 
       FUN = function(k) {
         n_distinct(mtcars[k])
       })
```


\

## Question 24

Get the first 9 observations of `mtcars`.

a.    Get the last 9 observations of `mtcars`

```{r}
# use tail()
tail(mtcars, 9)
```

b.    Get the 10th to 15th, 24th to 26th, and 28th to 32th observations of `mtcars`

```{r}
mtcars[c(10:15, 24:26, 28:32), ]
```

c.    Subset a data frame of the 15th to 25th observations of `hp`, `drat`, and `qsec`

```{r}
mtcars[15:25, c("hp", "drat", "qsec")]
```

d.    Subset the value of the 10th to 18th observations of `mpg` and `disp`

```{r}
# 10th to 18th observations of `mpg`
mtcars$mpg[10:18]

# 10th to 18th observations of `disp`
mtcars$disp[10:18]
```


\

## Question 25

Create a histogram of the `mpg` using `hist()` with appropriate x axis range and label the x axis

```{r}
hist(mtcars$mpg,
     breaks = 30, 
     xlim = c(10, 35),
     xlab = "mpg")
```


a.    Create a scatter plot of `qsec` versus `hp` by `carb`.

```{r}
library(ggplot2)

# scatter plot using ggplot
ggplot(mtcars) +
  aes(x = qsec, y = hp, colour = factor(carb)) +
  geom_point()
```




b.    Create a dot plot of `gear` variable by `drat`.

```{r}
# dot plots using ggplot2
ggplot(mtcars) +
  aes(x = factor(gear), y = drat) +
  geom_dotplot(binaxis = "y", stackdir = "center")
```

