---
title: "Exercise 1"
author: ""
date: "2023-04-07"
description: >
  Exercises for material for up to Module 3.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exercise 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Exercise 1

Install and load library `dplyr`.

a.  Call help for function arrange of `dplyr`

\

## Exercise 2

Create a variable called `myVar` and assign it to $30$

a.  What data type is `myVar`? (Use a function to check)
b.  Convert `myVar` to a different data type

\

## Exercise 3

Create a new variable called `my_num` tha contains 6 numbers

a.  Multiply `my_num` by 4

\

## Exercise 4

Create a second variable called `my_char` that contains 5 character strings

\

## Exercise 5

Combine the two variables that `my_num` and `my_char` into a variable called both

a.  What is the length of both
b.  Divide both by 3, what happens?

\

## Exercise 6

Create a vector with elements 1, 2, 3, 4, 5, 6 and call it `x`

\

## Exercise 7

Create another vector with elements 10, 20, 30, 40, 50 and call it `y`

\

## Exercise 8

What happens if you try to add `x` and `y` together? Why?

a.  Add `x` and `y` together
b.  Multiply `x` and `y` together. Pay attention to how R performs operations on vectors of the same length
c.  What happens if `x` and `y` have different length?

\

## Exercise 9

Now create a vector with elements "aa", "bb", "cc", "dd", "kk" and call it `z`

a.  What happens if you try to add `x` or `y` with `z` together
b.  Why does the error comes up? (Hint: data types)

\

## Exercise 10

If `x <- c("ww", "ee", "ff", "uu", "kk")`, what is the output for `x[c(2,3)]`?

a.  Can you subset `x` in another way and get the same output?

\

## Exercise 11

If `x <- c("ss", "aa", "ff", "kk", "bb")`, what will be the third value in the index vector operation `x[c(2,4,4)]`?

\

## Exercise 12

Assign `fname <-  "James"` and `lname <- "Bond"`,

a.  Produce the output `James Bond` using a line of code (Hint: use `paste()`)

\

## Exercise 13

Create two character vectors:

-   `Names <- c("John", "Andrew", "Thomas")` and
-   `Designation <- c("Manager", "Project Head", "Marketing Head")`.

a.  Write some R code to obtain the following output (Hint: check help page the function you used to create a data frame):

```{r echo=FALSE}
Names <- c("John", "Andrew", "Thomas")
Designation <- c("Manager", "Project Head", "Marketing Head")

names_design <- paste(Names, Designation)
df <- data.frame("Names Designation" = names_design, check.names = FALSE)
df
```

\

## Exercise 14

Create the following data frame, and call it Employees

```{r echo=FALSE}
Names <- c("Alex", "Lilly", "Mark", "Oliver", "Martha", "Lucas", "Caroline")
Age <- c(25, 31, 23, 52, 76, 49, 26)
Height <- c(177, 163, 190, 179, 163, 183, 164)
Weight <- c(57, 69, 83, 75, 70, 83, 53)
Sex <- c("F", "F", "M", "M", "F", "M", "F")

Employees <- data.frame(Names, Age, Height, Weight, Sex)
Employees
```

a.  Extract specific column from a data frame
b.  Extract 3rd and 5th rows with 1st and 3rd columns\
c.  Add a new column and call it Education, where everyone has an education status of HS
d.  Add new row called Louis
e.  Make `Sex` column into a factor. How many factor levels are in the columns?

\

## Exercise 15

Create the data frame below and call it Working (make sure you import the variable Working as character and not factor).

```{r echo=FALSE}
Working <- c("Yes", "No", "No", "Yes", "Yes", "No", "Yes")
df_3 <- data.frame(Names, Working)
df_3
```


a.  Add this data frame column-wise to the previous one.
b.  How many rows and columns does the new data frame have?
c.  What class of data is in each column?


\

# **Using the `iris` dataset in R**

## Exercise 16

Extract `Sepal.Length` from the `iris` dataset and call the resulting vector `mysepal`

a.  Get the summation, mean, median, max and min of `mysepal`
b.  Get the summary of `mysepal`

\

## Exercise 17

Get all rows of Species `versivcolor` in a new data frame. Call this data frame: `iris.vers`

\

## Exercise 18

Get a vector called `sepal.dif` with the difference between `Sepal.Length` and `Sepal.Width` of `versicolor` plants

\

## Exercise 19

Update (add) `iris.vers` with the new column `sepal.dif`

\

## Exercise 20

How many observations in `iris` gave a Sepal.Length greater or equal 5?

\

# **Using the `mtcars` dataset in R**

## Exercise 21

Check to see if you have the `mtcars` dataset by entering the command `mtcars`.

a.  What class is `mtcars`?

\

## Exercise 22

How many observations (rows) and variables (columns) are in the `mtcars` dataset?

\

## Exercise 23

Perform an descriptive analysis on the `mtcars` dataset (Hint: data types of columns, mean, median, ...).

a.    How many unique values in `cyl`, `vs`, `am`,  `gear`, and `carb` (Hint: use `unique()`)?

\

## Exercise 24

Get the first 9 observations of `mtcars`.

a.    Get the last 9 observations of `mtcars`
b.    Get the 10th to 15th, 24th to 26th, and 28th to 32th observations of `mtcars`
c.    Subset a data frame of the 15th to 25th observations of `hp`, `drat`, and `qsec`
d.    Subset the value of the 10th to 18th observations of `mpg`, `cyl`, and `disp`

\

## Exercise 25

Create s subset `cars` that only contains the columns: `wt`, `qsec`, and `hp` and assign this object to `carsSub`

a.  What are the dimensions of `carsSub`
b.  Extract the first 2 variables of the first 2 observations of `carsSub`
